---
title: Causal inference with potential outcomes bootcamp
subtitle: 'Part 2 of the GLM and causal inference series.'
author: A. Solomon Kurz
date: '2023-02-06'
tags:
  - ANCOVA
  - ANOVA
  - causal inference
  - g-computation
  - GLM
  - marginal standardization
  - potential outcomes
  - R
  - RCT
  - tidyverse
  - tutorial
draft: false
layout: single
featured: no
bibliography: /Users/solomonkurz/Dropbox/blogdown5/content/blog/my_blog.bib
biblio-style: apalike
csl: /Users/solomonkurz/Dropbox/blogdown5/content/blog/apa.csl  
link-citations: yes
---

One of the nice things about the simple OLS models we fit in the last post is they're easy to interpret. The various $\beta$ parameters were valid estimates of the population effects for one treatment group relative to the wait-list control. However, this nice property won't hold in many cases where the nature of our dependent variables and/or research design requires us to fit other kinds of models from the broader generalized linear mixed model (GLMM) framework. Another issue at stake is if you've spent most of your statistical analysis career in an OLS-state of mind, there's a good chance there are concepts that are undifferentiated in your mind. As is turns out, some of these concepts are important when we want to make valid causal inferences. Our task in this post is to start differentiating the undifferentiated.

## Reload and refit

### `horan1971` data.

In post, we'll be continuing on with our `horan1971` data set form the last post. These data, recall, were transposed from the values displayed in Table 2 from @horan1971coverant. I've saved them as an external `.rda` file in a `/data` subfolder on GitHub. If you don't want to wander over to my GitHub, you can just copy the code from my last post.

```{r, warning = F, message = F}
# packages
library(tidyverse)
library(marginaleffects)

# adjust the global theme
theme_set(theme_gray(base_size = 12) +
            theme(panel.grid = element_blank()))

# load the data
load(file = "data/horan1971.rda")

# wrangle a bit
horan1971 <- horan1971 %>% 
  filter(treatment %in% c("delayed", "experimental")) %>% 
  mutate(prec = pre - mean(pre),
         experimental = ifelse(treatment == "experimental", 1, 0))

# what are these, again?
glimpse(horan1971)
```

### Same old models.

In the last post, we explored how to fit an ANOVA- and ANCOVA-type model to the data using OLS regression. For this post, we'll refit those basic models. The ANOVA model for the data follows the form

$$
\begin{align*}
\text{post}_i & \sim \operatorname{Gaussian}(\mu_i, \sigma) \\
\mu_i & = \beta_0 + \beta_1 \text{experimental}_i,
\end{align*}
$$

where the two experimental conditions in play are captured by the dummy variable `experimental`. Here we fit that model again with the `lm()` function.

```{r}
# fit the ANOVA model
fit1 <- lm(
  data = horan1971,
  post ~ experimental
)

# summarize the results
summary(fit1)
```

We also learned the ANCOVA model for these data follows the form

$$
\begin{align*}
\text{post}_i & \sim \operatorname{Gaussian}(\mu_i, \sigma) \\
\mu_i & = \beta_0 + \beta_1 \text{experimental}_i + {\color{blueviolet}{\beta_2 \text{prec}_i}},
\end{align*}
$$

where $\beta_2$ is the coefficient for our baseline covariate `prec`, which is the mean-centered version of the participant weights (in pounds) before the intervention. Here's how to fit the ANCOVA-type model with `lm()`.

```{r}
# fit the ANCOVA model
fit2 <- lm(
  data = horan1971,
  post ~ 1 + experimental + prec
)

# summarize the results
summary(fit2)
```

As expected, the $\beta$ coefficients in the ANCOVA model all have smaller standard errors than those in the ANOVA model. Hurray! Statistics works!

## Causal inference

Okay, so at the beginning of the post, we said the $\beta$ coefficients for our experimental group are valid estimates of the population-level causal effects. But like, what does that even mean? Buckle up.

### Counterfactual interventions, no covariates.

Our friends in causal inference have been busy over the past few years. First, we should understand there are different ways of speaking about causal inference. I'm not going to cover all the various frameworks, here, but most of the causal inference textbooks I mentioned in the [last post](https://timely-flan-2986f4.netlify.app/blog/2023-02-06-boost-your-power-with-baseline-covariates/#i-make-assumptions) provide historical overviews. At the time of this writing, I'm a fan of the potential-outcomes framework [see @imbensCausalInferenceStatistics2015], the basics of which we might explain as follows: 

Say you have some population of interest, such as overweight female university students interested in losing weight [see @horan1971coverant]. You have some focal outcome variable $y$, which you'd like to see change in a positive direction. In our case that would be bodyweight, as measured in pounds. Since $y$ varies across $i$ persons, we can denote each participants' value as $y_i$. Now imagine you have 2 or more well-defined interventions. In our case, that would be assignment to the waitlist control or experimental intervention group. For notation purposes, we can let $0$ stand for the control group and $1$ stand for the active treatment group, just like with a dummy variable. We can then write $y_i^0$ for the $i$th person's outcome if they were in the control condition, and $y_i^1$ for the $i$th person's outcome if they were in the treatment condition. Putting all those pieces together, we can define the causal effect $\tau$ of treatment versus control for the $i$th person as

$$\tau_i = y_i^1 - y_i^0.$$

In the case of our `horan1971` data, the causal effect of the experimental treatment for each woman is her post-treatment weight for the experimental treatment minus her post-treatment weight for the waitlist condition. The problem, however, is that each woman was only randomized into one of the two conditions. And thus, each woman has, at best, only 50% of the data required to compute her individual causal effect, $\tau_i$. This is the so-called *fundamental problem of causal inference* [@holland1986statistics]; we are always missing at least half of the data. To help illustrate this, take a look at a random subset of the `horan1971`.

```{r}
set.seed(2)

horan1971 %>% 
  slice_sample(n = 10) %>% 
  mutate(y1 = ifelse(treatment == "experimental", post, NA),
         y0 = ifelse(treatment == "delayed", post, NA)) %>% 
  select(sn, treatment, post, y1, y0)
```

Within the `mutate()` function, I computed each participants' `y1` and `y0` score, based on a combination of her `treatment` and `post` values. Because none of the participants have values for both `y1` and `y0`, we cannot use the raw data to compute their individual treatment effects. What we can do, however, is compute the *average treatment effect* (ATE) with the formula:

$$\tau_\text{ATE} = \mathbb E (y_i^1 - y_i^0),$$

which, in words, just means that the average treatment effect in the population is the same as the average of each person's individual treatment effect. In the equation, I'm using the expectation operator $\mathbb E()$ to emphasize we're working within a likelihoodist framework. At first glance, it might appear[^1] this equation doesn't solve the problem that we cannot compute $y_i^1 - y_i^0$ from the data of any of our participants, because half of the required values are missing. However, it's also the case that when we're working with a simple OLS-type model[^2], 

$$\tau_\text{ATE} = \mathbb E (y_i^1 - y_i^0) = {\color{blueviolet}{\mathbb E (y^1) - \mathbb E (y^0)}},$$

where $\mathbb E (y^1)$ is the population average of our $y_i^1$ values, and $\mathbb E (y^0)$ is the population average of our $y_i^0$ values. Sometimes causal inference scholars differentiate between the *sample* average treatment effect (SATE) and the *population* average treatment effect (PATE). In this blog post and in the rest of this series, I'm presuming researchers are analyzing their data with regression models to make population-level inferences. Thus, I'm always equating the ATE with the PATE. While I'm at it, there are other technical caveats which have to do with proper randomization and whether participants are truly independent of one another and so on. For the sake of this series, I'm presuming a clean simple randomization with no further complications. If you want more complications, check out @imbensCausalInferenceStatistics2015 and any of the other texts on causal inference. Trust me, we already have enough complications on our hands.

### Estimands, estimators, and estimates.

Before we get into it, we should probably introduce a few more vocabulary words. 

* An **estimand** is the focal quantity of interest. It's the reason we're analyzing our data and it's the answer to our primary research question. From a causal inference perspective, the estimand is the population-level causal effect.
* **Estimators** are the statistical methods we use to analyze our data. In this blog post and in the last, our estimators have been our OLS regression models. In the next couple blog posts, we'll add logistic regression via maximum likelihood and various Bayesian models to our list of estimators.
* An **estimate** is the result from your statistical model (estimator) that's designed to answer your research question.

If you prefer your education in the form of silly social media memes, maybe this one will help:

```{r echo = FALSE}
blogdown::shortcode('tweet', user = "rlmcelreath", id = '1582368904529137672')
```

Anyway, next we'll learn how to actually compute $\tau_\text{ATE}$ within the context of our OLS models. This will be the be the estimate of our estimand.

#### $\mathbb E (y^1) - \mathbb E (y^0)$ from `fit1`.

Sometimes the authors of introductory causal inference textbooks have readers practice computing these values by hand, which can have its pedagogical value. But in your role as a professional scientist, you'll be computing $\tau_\text{ATE}$ within the context of a regression model, so you can properly express the uncertainty of your estimand with 95% intervals, as standard error, or some other measure of uncertainty. With our simple `fit1` model, we can compute $\mathbb E (y^1)$ and $\mathbb E (y^0)$ with the base **R** `predict()` function.

```{r}
nd <- tibble(experimental = 0:1)

predict(fit1, 
        newdata = nd,
        se.fit = TRUE,
        interval = "confidence") %>% 
  data.frame() %>% 
  bind_cols(nd)
```

The `fit.fit` column shows the point estimates, and the `fit.lwr` and `fit.upr` columns show the 95% intervals, and the `se.fit` columns shows the standard errors. Though the `predict()` method is great for computing $\mathbb{E}(y^1)$ and $\mathbb{E}(y^0)$, it doesn't give us a good way to compute the difference of those values with a measure of uncertainty, such as a standard error. Happily, we can rely on functions from the handy **marginaleffects** package [@R-marginaleffects] for that. First, notice how the `predictions()` function works in a similar way to the `predict()` function from above, but with nicer default behavior.

```{r}
predictions(fit1, newdata = nd, by = "experimental")
```

The **marginaleffects** package offers a few ways to contrast the to mean values. With our current `predictions()` workflow, we can just add in `hypothesis = "revpairwise"`.

```{r}
predictions(fit1, newdata = nd, by = "experimental", hypothesis = "revpairwise")
```

Now we have a nice standard error and 95% interval range for the estiamte of our estimand, $\tau_\text{ATE}$. Thus, the average causal effect of the experimental condition relative to the waitlist control is a reduction of about 2 and a half pounds, with with a very wide 95% confidence interval range spanning from a reduction of 13 pounds to an *increase* of 8 pounds. With that all in mind, look back at the parameter summary for `fit1`.

```{r}
summary(fit1)
```

Notice that the summary for our $\beta_1$ parameter is the same as the $\tau_\text{ATE}$ from above. When you have a simple OLS-type Gaussian model without a fancy link function, the $\tau_\text{ATE}$ will be the same as the $\beta$ coefficient for the treatment dummy.

#### $\mathbb E (y_i^1 - y_i^0)$ from `fit1`.

It's time for me to confess my rhetoric above was a little misleading. As it turns out, you can in fact compute $\mathbb E (y_i^1 - y_i^0)$ from your regression models, even with 50% of the values missing. The key is to compute *counterfactual* values for $y_i^1$ and $y_i^0$ from the model. Before we can do that, we'll first need to redefine our `nd` predictor data.

```{r}
nd <- horan1971 %>% 
  select(sn) %>% 
  expand_grid(experimental = 0:1)

# what?
glimpse(nd)
```

What we've done is taken each unique case in the original `horan1971` data, as indexed by `sn`, and assigned them both values for the `experimental` dummy, `0` and `1`. As a consequence, we took our 41-row data frame and doubled it in length. Now we can insert our updated counterfactual `nd` into the base **R** `predict()` to compute all those $y_i^1$ and $y_i^0$ estimates.

```{r}
predict(fit1, 
        newdata = nd,
        se.fit = TRUE,
        interval = "confidence") %>% 
  data.frame() %>% 
  bind_cols(nd) %>% 
  # just show the first 6 rows
  head()
```

Now each case (`sn`) gets their own estimate for both levels of the `experimental` dummy. Given these are counterfactual estimates from a statistical model, they also come with their measures of uncertainty. But just like before, the `predict()` method doesn't give us a good way to use these model predictions to compute $\mathbb E (y_i^1 - y_i^0)$ in a way that accounts for the standard errors. Once again, the **marginaleffects** package has the solution. Like before, our first attempt will be to insert our updated `nd` data into the `predictions()` function. This time, we've included both the `sn` and `experimental` variables into the `by` argument, to help clarity the output.

```{r}
predictions(fit1, newdata = nd, by = c("sn", "experimental")) %>% 
  head()
```

I've used the `head()` function to limit the output to the first six rows, but the full output would have all 82 rows worth of counterfactual predictions. Each one has its own standard error and so on. To compute the actual participant-level contrasts, $y_i^1 - y_i^0$, we'll want to switch to the `marginaleffects::comparisons()` function. Here we just need to use the `variables` function to indicate we want counterfactual comparisons on the `experimental` dummy for each case in the original data set.

```{r}
comparisons(fit1, variables = list(experimental = 0:1)) %>% 
  head()
```

Here we see the $y_i^1 - y_i^0$ for the first six participants in the data set, each with its own standard errors and so on. But recall our estimand $\tau_\text{ATE}$ is defined as $\mathbb E (y_i^1 - y_i^0)$. This means, we need a way to compute the average of those contrasts, with a method that also accounts for their standard errors. Happily, all we need to do is tack on the `summary()` function, which will prompt the **marginaleffects** to compute the average of those participant-level contrasts and use the so-called delta method to compute the accompanying standard error.

```{r}
comparisons(fit1, variables = list(experimental = 0:1)) %>% 
  summary()
```

Again in this case our model-based estimate for $\tau_\text{ATE}$, computed by the formula $\mathbb E (y_i^1 - y_i^0)$, is the same as the $\beta_1$ coefficient and its standard error. So in the case of a simple OLS-based ANOVA-type model of a randomized experiment,

* $\beta_1$,
* $\mathbb E (y_i^1 - y_i^0)$, and
* $\mathbb E (y^1) - \mathbb E (y^0)$

are all the same thing. They're all estimators of our estimand $\tau_\text{ATE}$, the average treatment effect.

### Counterfactual interventions, with covariates.

Much like we can use baseline covariates when we analyze RCT data to boost the power, we can use baseline covariates when we make causal inferences, too. We just have to expand the framework a bit. If we let $c_i$ stand for the $i$th person's value on continuous covariate $c$, we can define the ATE, conditional on covariates $c$, with the formula:

$$\tau_{\text{ATE}|c} = \mathbb E (y_i^1 - y_i^0 | c_i),$$

which, in words, means that the average treatment effect in the population, computed conditional on $c$, is the same as the average of each person's individual treatment effect, computed conditional on their values of $c$. We can generalize this equation so that $\mathbf C_i$ is a vector of covariates, continuous or otherwise, the values for which vary across the participants, to the following:

$$\tau_{\text{ATE}|\mathbf C} = \mathbb E (y_i^1 - y_i^0 | \mathbf C_i).$$

Though we won't consider more complex data examples in this blog post, we will want to keep the $\mathbf C$ vector insights in the backs of our minds for the blog posts to come. In the literature, this method is often called *standardization* or *g-computation*. To my knowledge, these terms have their origins in different parts of the literature, but they're really the same thing when used in the ways I'll be highlighting in this blog series. For a way into this literature, you might check out @snowden2011implementation, @muller2014estimating, or @wang2017gcomputation.

Anyway, an alternative approach is to use the mean[^3] value of the covariate, $\bar c$, to compute the conditional predicted values for the two levels of treatment, and then take their difference:

$$\tau_{\text{ATE}| \bar c} = \mathbb E (y^1 | \bar c) - \mathbb E (y^0 | \bar c).$$

As above, we can generalize this equation so that $\mathbf C_i$ is a vector of covariates, and define the equation for the ATE conditional on $\mathbf{\bar C}$ to

$$\tau_{\text{ATE}| \mathbf{\bar C}} = \mathbb E \left (y^1 | \mathbf{\bar C} \right) - \mathbb E \left (y^0 | \mathbf{\bar C} \right).$$

As we will see, when working with a continuous outcome variable within the conventional OLS-type paradigm, 

$$\mathbb E (y_i^1 - y_i^0 | c_i) = \mathbb E (y^1 | \bar c) - \mathbb E (y^0 | \bar c),$$

and

$$\mathbb E (y_i^1 - y_i^0 | \mathbf C_i) = \mathbb E \left (y^1 | \mathbf{\bar C} \right) - \mathbb E \left (y^0 | \mathbf{\bar C} \right).$$

Here's what this looks like in action:

#### $\mathbb E (y^1 | \bar c) - \mathbb E (y^0 | \bar c)$ from `fit2`.

With our ANCOVA-type `fit2` model, we can compute $\mathbb E (y^1 | \bar c)$ and $\mathbb E (y^0 | \bar c)$ with the base **R** `predict()` function. As a first step, we'll define our prediction grid with the sample means for our covariate `prec`, and then expand the grid to include both values of the `experimental` dummy.

```{r}
nd <- horan1971 %>% 
  summarise(prec = mean(prec)) %>% 
  expand_grid(experimental = 0:1)

# what?
print(nd)
```

Since the `prec` covariate was already mean centered, we technically didn't need to manually compute `mean(prec)`. We already knew that value would be zero. But I wanted to make the point explicit so the step will generalize to other data contexts. Anyway, now we have our `nd` data, we're ready to pump the values into `predict()`.

```{r}
predict(fit2, 
        newdata = nd,
        se.fit = TRUE,
        interval = "confidence") %>% 
  data.frame() %>% 
  bind_cols(nd)
```

Similar to the simple ANOVA-type `fit1` version fo the model, the `predict()` method is great for computing $\mathbb{E}(y^1 | \bar c)$ and $\mathbb{E}(y^0 | \bar c)$, but it doesn't give us a good way to compute the difference of those values with a measure of uncertainty. For that, we can once again rely on the `marginaleffects::predictions()` function.

```{r}
predictions(fit2, newdata = nd, by = "experimental")
```

To get the contrast, just add in `hypothesis = "revpairwise"`.

```{r}
predictions(fit2, newdata = nd, by = "experimental", hypothesis = "revpairwise")
```

And also like with the ANOVA-type `fit1`, this method for computing the $\tau_{\text{ATE}| \bar c}$ from `fit2` returns the same estimate and uncertainty statistics as returned by the `summary()` information for the $\beta_1$ parameter.

```{r}
summary(fit2)
```

When you fit an OLS-type ANCOVA model with the conventional identity link, $\mathbb{E}(y^1 | \bar c) - \mathbb{E}(y^0 | \bar c)$ will be the same as the $\beta$ coefficient for the treatment dummy. They're both estimators of the estimand $\tau_{\text{ATE}| \bar c}$.

#### $\mathbb E (y_i^1 - y_i^0 | c_i)$ from `fit2`.

Before we compute our counterfactual $\mathbb{E}(y_i^1 - y_i^0 | c_i)$ estimates from our ANCOVA-type `fit2`, we'll first need to redefine our `nd` predictor data. This time, we'll retain each participants' `prec` value (i.e., $c_i$).

```{r}
nd <- horan1971 %>% 
  select(sn, prec) %>% 
  expand_grid(experimental = 0:1)

# what?
glimpse(nd)
```

Now each level of `sn` has two rows, one for each of the `experimental` dummy's values: `0` and `1`. But within each level of `sn`, the baseline covariate `prec` is held constant to its original value. Now we can insert our updated counterfactual `nd` into the base **R** `predict()` to compute the conditional estimates for `post`.

```{r}
predict(fit2, 
        newdata = nd,
        se.fit = TRUE,
        interval = "confidence") %>% 
  data.frame() %>% 
  bind_cols(nd) %>% 
  # subset the output
  head()
```

With a little more wrangling, we can compute the point estimates for $(y_i^1 - y_i^0 | c_i)$.

```{r}
predict(fit2, 
        newdata = nd,
        se.fit = TRUE,
        interval = "confidence") %>% 
  data.frame() %>% 
  bind_cols(nd) %>% 
  select(sn:experimental, fit.fit) %>% 
  pivot_wider(names_from = experimental, values_from = fit.fit) %>% 
  mutate(tau = `1` - `0`)
```

Even though the participants vary on their point estimates for `0` and `1`, they all have the same estimates for their difference, `tau`. This is a normal characteristic of analyses within the OLS-type paradigm, but it will not hold once we generalize to other kinds of likelihoods. You'll see. But anyways, since this workflow won't allow us to retain the uncertainty statistics, we'll switch back to our **marginaleffects**-based workflow. As a first step, we insert our updated `nd` data into the `predictions()` function. This time we include the `sn`, `experimental`, and `prec` variables into the `by` argument, to help clarity the output.

```{r}
predictions(fit2, newdata = nd, by = c("sn", "experimental", "prec")) %>% 
  head()
```

The `head()` function to limited the output to the first six rows, but the full output would have all 82 rows worth of counterfactual predictions. Each one has its own standard error and so on. To compute the actual participant-level contrasts, $(y_i^1 - y_i^0 | c_i)$, we switch to the `marginaleffects::comparisons()` function. Here we just need to use the `variables` argument to indicate we want counterfactual comparisons on the `experimental` dummy for each case in the original data set.

```{r}
comparisons(fit2, variables = list(experimental = 0:1)) %>% 
  head()
```


Here we see the $(y_i^1 - y_i^0 | c_i)$ for the first six participants in the data set, each with its own standard errors and so on. But recall our estimand $\tau_{\text{ATE}|c}$ is estimated via $\mathbb E (y_i^1 - y_i^0 | c_i)$, which means we need to compute the average of those contrasts in a way that produces a standard error. As with the simpler ANOVA-type workflow we used with `fit1`, we can simply tack on a `summary()` line, which will compute delta-method based standard errors.

```{r}
comparisons(fit2, variables = list(experimental = 0:1)) %>% 
  summary()
```

When our goal is just to compute $\mathbb E (y_i^1 - y_i^0 | c_i)$, we can also use the `marginaleffects::avg_comparisons()` function, which skips the `summary()` step.

```{r}
avg_comparisons(fit2, variables = list(experimental = 0:1))
```

As Arel-Bundock pointed out his [-@arelBundock2023CausalInference] vignette, [*Causal inference with the parametric g-formula*](https://vincentarelbundock.github.io/marginaleffects/articles/gformula.html), the `avg_comparisons()` function is a compact way to compute $\mathbb E (y_i^1 - y_i^0 | c_i)$ with the parametric g-formula method. Again in this case our estimate for $\tau_{\text{ATE}|c}$ via $\mathbb E (y_i^1 - y_i^0 | c_i)$ is the same as the $\beta_1$ coefficient and its standard error from `fit2`, and they're both the same as $\tau_{\text{ATE}| \bar c}$ via $\mathbb E (y^1 | \bar c) - \mathbb E (y^0 | \bar c)$. We might further say that, in the case of an OLS-based ANCOVA-type model of a randomized experiment,

* $\beta_1$,
* $\mathbb{E}(y_i^1 - y_i^0 | c_i)$, and
* $\mathbb{E}(y^1 | \bar c) - \mathbb{E}(y^0 | \bar c)$

are all the same thing. They're all estimators of our estimand, the average treatment effect. We can extend this further to point out that the three estimators we used with our ANOVA-type model `fit1` were also estimators of the average treatment effect. But the three methods we just used for our ANCOVA-type model `fit2` all benefit from the increased precision (i.e., power) that comes from including a high-quality baseline covariate in the model.

## Recap

In this post, some of the main points we covered were:

* The potential-outcomes framework is one of the contemporary approaches to causal inference.
* We cannot compute an individual's causal effect $\tau_i = y_i^1 - y_i^0$ by hand, because we are always missing at least half of the data. This is the *fundamental problem of causal inference*.
* Conceptually, the *average treatment effect* (ATE, $\tau_\text{ATE}$) is the mean of the person-specific treatment effects.
* In a simple ANOVA-type regression model, we can estimate $\tau_\text{ATE}$ with either the $\mathbb E (y_i^1 - y_i^0)$ or the $\mathbb E (y^1) - \mathbb E (y^0)$ method, and the results will be exactly the same. Both methods will also be the same as the $\beta$ coefficient for the treatment dummy in the ANOVA model.
* With an ANCOVA-type regression model with a single baseline covariate, we can estimate $\tau_\text{ATE}$ with either the $\mathbb E (y_i^1 - y_i^0 | c_i)$ or the $\mathbb{E}(y^1 | \bar c) - \mathbb{E}(y^0 | \bar c)$ method, and the results will be exactly the same, granted we use the same covariate $c$.
* We can generalize the two ANCOVA-type methods to models with multiple baseline covariates.

At this point, some readers might wonder why we have so many methods that produce the identical results. As we will soon see, this pattern will not generalize to models with other likelihoods and link functions. Speaking of which, in the next post we'll see what this framework looks like for logistic regression.

See you in the next one, friends!

## Session information

```{r}
sessionInfo()
```

## References

[^1]: I say "might appear" because we'll eventually see this isn't an unsolvable problem within our regression paradigm. But we do have to make some strong assumptions with our paradigm and the counterfactual estimates we'll produce still aren't the same thing as if we had actually observed all the potential outcomes. But we're getting ahead of ourselves.

[^2]: As we'll see later, this has a lot to do with link functions, which are popular with non-Gaussian models like logistic and Poisson regression. But basically, if you're fitting a mean-model with an identity link, as with conventional OLS or a simple Gaussian GLM, $\mathbb E (y_i^1 - y_i^0) = \mathbb E (y^1) - \mathbb E (y^0)$. In other situations, it might not.

[^3]: In addition to the sample mean, you could also use the mode [see @muller2014estimating], or some other value of theoretical interest. The mode can be a particularly good option when working with categorical covariates, and we'll cover this possibility in a future post. If you have several variables in your covariate set ($\mathbf{C}$), you could even use different strategies for different covariates.

```{r, eval = F, echo = F}
tibble(x = 0, y = 0, label = "tau[ATE]") %>% 
  ggplot(aes(x = x, y = y, label = label)) +
  geom_text(parse = T, family = "serif", size = 20) +
  theme_void() +
  theme(plot.background = element_rect(color = "white", fill = "#fff7cc"))

ggsave("tau-ate-featured.jpg", width = 3.5, height = 3.5, units = "in")
```

